Stochastic Wind Power Forecasting

### 2020-06-09 LaTeX_Code/Paper/paper_1line_Renzo.tex
Modifications in paper. Still there are more to do (hopefully I will finish today)

Reliable wind power generation forecasting is crucial for applications such as the allocation of energy reserves, optimization of electricity price and operation scheduling of conventional power plant. We propose a data driven model based on parametric Stochastic Differential Equations (SDEs) to captures real-world asymmetric dynamics of wind power forecast errors. Our SDE framework incorporates time derivative tracking of the forecast, time-dependent mean reversion parameter and an improved diffusion term. We are able to simulate future wind power production paths and to get sharp confidence bands. The method is forecast technology agnostic and enables the comparison between different forecasting technologies on the basis of information criteria. We apply the model to historical Uruguayan wind power production data and forecasts on the year 2019.

@@

{\color{red}Keywords: Indirect inference, wind power, probabilistic forecasting, stochastic differential equations, Lamperti transform, model selection.}

{\color{red} 2020 AMS Subject Classification: 60H10}

@@

Different methods have been applied to wind power forecasting. They can be generally categorized as follows: physical models, statistical methods, artificial intelligence methods and hybrid approaches. The output of such methods is usually a deterministic forecast. Occasionally probabilistic forecasts are produced through uncertainty propagation in the data, parameters or through forecast ensembles. {\color{red} Expand discussion about works on probabilistic forecasting.} However, there is a lacking in simulating and producing data driven stochastic forecasts based on real-world performance of forecasting models. It is crucial to capture the forecast's actual performance as it has been known that different forecasting technologies exhibit different behavior for different wind farms and seasons (\cite{chang}). This is due to many factors which forecast are challenged to capture such as the surrounding terrains of the wind farm and the condition of the blades such as icing, wear and tear or dirt. It is known that complex terrains in both off shore and on shore locations decrease the accuracy of wind power forecasts significantly (\cite{schicker2017short}). It also has been shown that the performance of forecasts varies from month to month. Thus the performance of wind power forecasts is location and time dependent.

Many approaches have been taken to evaluate the uncertainty of a given forecast. There are two types of errors: level errors and phase errors. The use of mean or median errors in this context may be misleading as wind power forecast errors are asymmetric. This is a natural consequence of wind power being non-negative and bounded by the maximum capacity of production. This is important as the associated cost to power forecast errors are also asymmetric due to different costs for up and down power regulations which are determined by the electricity market (\cite{tsitsiklis2015pricing}).

We propose to model wind power forecasts errors using parametric stochastic differential equations (SDEs) whose solution defines a stochastic process. This resultant stochastic process describes the time evolution dynamics of wind power forecast errors while capturing properties such as a correlation structure and the inherent asymmetry. Additionally, the model we propose is agnostic of the forecasting technology and serves to complement forecasting procedures by providing a data driven stochastic forecast. Hence, we are able to evaluate wind power forecasts according to their real-world performance and we are able to compare different forecasting technologies. Most notably, we are able to simulate future wind power production given a deterministic wind power forecast. Future wind power production using Monte Carlo methods, as well as the analytic form of the proposed SDE, can be used in optimal control problems involving wind power production.

Previous attempt by (\cite{mozuma}) considered stochastic wind power forecast models based on stochastic differential equations. Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion and non-Gaussian approximations. We apply the model to Uruguayan wind power forecasts together with historical wind power production data pertaining to the year 2019. \\

In Section \ref{Section_2}, we introduce the main characteristics of a real data set encompassing the normalized wind power production in Uruguay during the period April-December 2019, joint with the most accurate predictions, as highlighted in our posterior analysis, performed by one out of the three sources of forecast providers. The significant steps for constructing phenomenological models of the normalized wind power production and the forecast error based on stochastic differential equations are described in Section \ref{Section_3}. After reformulating the SDE for the forecast error in Section \ref{Section_4}, by applying the Lamperti transform that leads to a unit diffusion coefficient, we write down in Section \ref{Section_5} the expressions for the likelihood functions of the forecast error in its original space and in the Lamperti space. We also derive simple approximations of the likelihood functions. Section \ref{Section_5} concludes with the description of the optimization algorithm to compute approximate maximum likelihood estimates, including the case where we expand the model comprising an initial transition from the time the forecast is performed to the time of the first forecast.  
In Section \ref{Section_6}, we apply our proposed parametric estimation procedure to the Uruguay wind and forecast dataset, comparing two alternative models and assessing, for the best candidate model, the performance of three different forecast providers.
The proofs of the existence, strong uniqueness, and boundedness of the solutions of the SDE used to model normalized wind power production and its forecast error are given in the Appendix.

@@

A view of the global discrepancy between the real production and the forecasted production, during the nine months observation period, is summarized through the forecast error histograms, Figure \ref{fig:data_curtailing}, where we also partitioned the forecast errors according to three contiguous categories of normalized generated power. Low normalized generated power corresponds to the range $[0,0.3]$, mid-power refers to the range $(0.3,0.6]$, and high-power to the range $(0.6,1]$.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_6.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_6.eps}
\caption{Wind production forecast error histograms during the period April-December 2019: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
  \label{fig:data_curtailing}
\end{figure}

We may observe that all the histograms in the Figure \ref{fig:data_curtailing} exhibit skewed patterns, to a different extent, as well as extreme observations. The presence of these features can be partly explained. The analysis of the dataset highlighted that, during several 24-hour segments, the system operators decided to reduce, or even cease the wind power production. Indeed, as recalled in (\cite{irena2}, p.3), ``Uruguay experiences high curtailment levels because generation exceeds demand.'' Despite the large country’s interconnection capacity with Argentina and Brazil, there is no active cross-border market, the energy is traded via ad hoc short-term agreements. ``Even with interconnection capacity exceeding peak demand, the power system experiences high VRE curtailment, mostly at night when wind generation exceeds demand.''

The curtailment of the wind power production imposed by the system operators has a strong influence on the forecast error. To build a model that, driven by the available forecast, allows to include the true power production with a prescribed degree of uncertainty, it is necessary to remove the data segments affected by wind curtailment.

@@

Let $(p_t)_{t \in [0,T]}$ be the available prediction function for the normalized wind power, which is an input to this approach. To start with the model specification, first we introduce a time-dependent drift function that features the mean-reverting property as well as derivative tracking:
\begin{equation}
a(X_t; p_t, \dot{p}_t, \bm{\theta}) = \dot{p}_t  - \theta_t (X_t - p_t)\,,  \label{drift:meanrev-derivtrack}
\end{equation} 
where $ (\theta_t)_{t \in [0,T]} $ is a positive deterministic function, whose range depends on $\bm{\theta}$, as will be explained shortly.

Now, looking at the normalized wind power generation forecast process $X$, modeled as solution to the It\^{o} stochastic differential equation (\ref{eq:main}) with the drift specified in (\ref{drift:meanrev-derivtrack}), it is straightforward to check that $\mathbb{E} \left[X_t\right] = p_t$, given $  \mathbb{E}\left[X_0\right] = p_0$. The application of  It\^{o}'s lemma on $g(X_t, t) = X_t e^{\int_{0}^{t} \theta_s \dif s}$, leads to 
\begin{equation*}
\dif \left( X_t e^{\int_{0}^{t} \theta_s \dif s} \right) = e^{\int_{0}^{t} \theta_s \dif s} (\dot{p}_t + \theta_t p_t ) \dif t  + e^{\int_{0}^{t} \theta_s \dif s}b (X_t; p_t, \dot{p}_t, \bm{\theta} ) \dif W_t,
\end{equation*}
whose integral form is
\begin{equation}
e^{\int_{0}^{t} \theta_s \dif s}  X_t - X_0 = \int_{0}^{t}  (\dot{p}_s + \theta_s p_s ) e^{\int_{0}^{s} \theta_u \dif u} \dif s + \int_{0}^{t}  b (X_s; p_s, \dot{p}_s, \bm{\theta} )  e^{\int_{0}^{s} \theta_u \dif u}\dif  W_s.   \label{eq:tomeanX}
\end{equation} 
Taking expectation on (\ref{eq:tomeanX}) we obtain
\begin{align}
\mathbb{E} \left[X_t\right] & = e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E} \left[X_0\right] +   \int_{0}^{t}  (\dot{p}_s + \theta_s p_s ) e^{\int_{0}^{s} \theta_u \dif u} \dif s   \right) \nonumber \\ 
& =  e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E} \left[X_0\right] +   \int_{0}^{t}  \dot{p}_s e^{\int_{0}^{s} \theta_u \dif u} \dif s +  \big[ p_s \, e^{\int_{0}^{s} \theta_u \dif u} \big]^{t}_0  - \int_{0}^{t}  \dot{p}_s e^{\int_{0}^{s} \theta_u \dif u} \dif s  \right) \nonumber \\
& =  e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E}\left[ X_0\right] + p_t \, e^{ \int_{0}^{t} \theta_s \dif s} - p_0 \right) = p_t. \label{eq:meanX}
\end{align}

 At this stage, the process defined by (\ref{eq:main}) with drift (\ref{drift:meanrev-derivtrack}) satisfies the two following properties: 
\begin{itemize}
\item it reverts to its mean $p_t$, with a time-varying speed $ \theta_t$ that is proportional to the deviation of the process $X$ from its mean,
\item it tracks the time derivative $\dot{p}_t$.  
\end{itemize}

@@

\begin{figure}[H]
\centering
  \includegraphics[width=0.45\textwidth]{plots/31.eps}\includegraphics[width=0.45\textwidth]{plots/31_2.eps}
  \caption{Pointwise confidence bands fitted, for the same daily segment, through diffusion models without derivative tracking (graph on the left) and with derivative tracking (graph on the right).}
  \label{fig:derivative_tracking}
\end{figure}

The forecast and production wind power data of Uruguay are normalized with respect to the installed power capacity during the period of observation. Thus, the mean-reverting level lies in $[0,1]$, and the process $X$  must take values in the same interval, a requirement that is not automatically fulfilled through the derivative tracking. To impose that the state space of $X$ is $[0,1]$, we may choose a convenient diffusion term, and require that the time-varying parameter $ \theta_t$ satisfies an ad-hoc condition.

Let $\bm{\theta} = (\theta_0,\alpha)$, and choose a state-dependent diffusion term that avoids the process exiting from the range $[0,1]$ as follows:
  \begin{equation}
    b (X_t; \bm{\theta} )= \sqrt{2 \alpha \theta_0 X_t (1 - X_t)}
  \end{equation}
  where $\alpha >0$ is a unknown parameter that controls the path variability. This diffusion term belongs to the Pearson diffusion family and, in particular, it defines a Jacobi type diffusion. It is useful to recall that (\cite[440]{foso}) a Pearson diffusion is a stationary solution to a stochastic differential equation of the form
 \begin{equation}
    \dif X_t = - \theta (X_t - \mu)\dif t + \sqrt{2 \theta \left(a X_t^2 + b X_t + c\right)} \dif W_t
  \end{equation}
where $\theta>0$, and $a$, $b$ and $c$ are parameters such that the square root is well defined when $X_t$ is in the state space. These parameters, together with $\mu$, the mean of the invariant distribution, determine the state space of the diffusion as well as the shape of the invariant distribution.

@@

Our model (\ref{VtSDE}) for the forecast error has a diffusion term that depends on the state variable $V_t$. Under the conditions that permit to use It\^{o}'s formula on a well-chosen transformation of the process $V$, John Lamperti (\cite{lamp}) first showed that the transformed process is again a diffusion that is solution to a SDE with unit coefficient for the diffusion term. 
A vast literature refers nowadays to this result as the so-called Lamperti transform (see, for example, \cite[40--41]{iacus1}, \cite{moma}, \cite[199--203]{pani}, \cite[98--100]{saso}), which is a basic tool to obtain a SDE for the transformed process whose diffusion term does not depend anymore on the state variable. A remarkable effect of removing the state dependency from the random noise term is to increase the numerical stability of the simulated paths of the transformed process. 
For this reason, some estimation methods of the unknown parameters of non-linear SDE models incorporated the use on the Lamperti's change of variable as part of a more complex approximation procedure (for example, the local linearization method in  \cite{shoz}, or the expansion method in \cite{ait}, later extended to multivariate diffusion in \cite{ait2}).

We consider the following Lamperti transformation  
%{\color{red} (Remind that continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
%\begin{equation}
%Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
%\end{equation}
%where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, 
\begin{align}
Z_t = h(V_t, t; \bm{\theta} )  & = \frac{1}{\sqrt{2 \alpha \theta_0 }} \int \frac{1}{\sqrt{(v + p_t)(1 - v - p_t)}} \dif v \Bigg{\vert}_{v = V_t}   = - \sqrt{ \frac{2}{ \alpha \theta_0 }} \arcsin (\sqrt{ 1 - V_t - p_t}) \label{eq:LampZ}
\end{align}
that, after applying It\^{o}'s formula on $h(V_t, t)$, leads to the following SDE with state independent unit diffusion term
\begin{align}
\dif Z_t = \Bigg[  & \frac{\dot{p}_t}{ \sqrt{2 \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}}  + \frac{- \theta_t V_t}{ \sqrt{2  \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \alpha \theta_0} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \Bigg] \dif t + \dif W_t \,. \label{eq:stindepSDE}
\end{align}
After replacing $V_t = 1 - p_t - \sin^2 \left(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \right) $ in (\ref{eq:stindepSDE}), we obtain that the  process $Z$ satisfies the SDE
\begin{align}
\dif Z_t & = \left[ \frac{ \dot{p}_t  - \theta_t  \left(1 - p_t - \sin^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\sqrt{2 \alpha \theta_0} \cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)}   - \frac{1}{4}  \frac{\sqrt{2 \alpha \theta_0} \left(1 - 2 \cos^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)} \right]\dif t + \dif W_t \nonumber \\
&  = \left[  \frac{  2  \dot{p}_t - \theta_t (1 - 2 p_t)  + (\alpha \theta_0 - \theta_t) \cos(- \sqrt{2 \alpha \theta_0 } Z_t) }{\sqrt{2 \alpha \theta_0} \sin{(- \sqrt{2 \alpha \theta_0} Z_t)}}  \right] \dif t + \dif W_t.  \label{eq:stindepSDE2}
\end{align}

We can see in Figure (\ref{fig:LP_transitions}) how the Lamperti transformation modifies the forecast error transition histogram without curtailment displayed in Figure (\ref{fig:error_transitions}).

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t_LP.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t_LP.eps}
\caption{Lamperti transformed forecast error transition histograms during the period April-December 2019 without wind power production curtailment: low-power (upper-left graph), mid-power (upper-right graph), high-power (lower-left graph), and the global range of power (lower-right graph).}
  \label{fig:LP_transitions}
\end{figure}

The shape of the forecast error transition histograms after the Lamperti transformation has much resemblance with a Gaussian distribution, motivating toward the use of Gaussian-like approximations of the unknown density transition functions of the process Z.

@@

\item Use the training dataset to find the SDE's first and second moments as explained in subsection \ref{moments_ODEs}.
\item Match the proxy distribution moments with the SDE's moments.
\item Evaluate the negative log-likelihood using the training dataset.

>>>>

Reliable wind power generation forecasting is crucial for applications such as the allocation of energy reserves, optimization for electricity price, and operation scheduling of conventional power plants. We propose a data-driven model based on parametric Stochastic Differential Equations (SDEs) to capture the real asymmetric dynamics of wind power forecast errors. Our SDE framework features time-derivative tracking of the forecast, time-varying mean-reversion parameter, and an improved state-dependent diffusion term. The methodology we developed allows the simulation of future wind power production paths and to obtain sharp empirical confidence bands. All the procedure is agnostic of the forecasting technology and enables comparing different forecast providers. We apply the model to historical Uruguayan wind power production data and forecasts during the period April-December 2019.

@@

Keywords: Wind power, probabilistic forecasting, stochastic differential equations, Lamperti transform,

numerical optimizaton, model selection.

2020 AMS Subject Classification: 60H10, 62M20, 65K10.

@@

Different methods have been applied to wind power forecasting. They can be generally categorized as follows: physical models, statistical methods, artificial intelligence methods, and hybrid approaches. The output of such methods is usually a deterministic forecast. Occasionally probabilistic forecasts are produced through uncertainty propagation in the data, parameters, or forecast ensembles. {\color{red} Expand discussion about works on probabilistic forecasting.} However, there is a lack of simulating and producing data-driven stochastic forecasts based on forecasting models. It is crucial to capture the forecast's actual performance as it has been known that different forecasting technologies exhibit different behavior for different wind farms and seasons (\cite{chang}). This is due to many factors that forecast are challenged to capture, such as the surrounding terrains of the wind farm and the condition of the blades such as icing, wear and tear, or dirt. It is known that complex terrains in both off shore and on shore locations decrease the accuracy of wind power forecasts significantly (\cite{schicker2017short}). It also has been shown that the performance of forecasts varies from month to month. Thus the performance of wind power forecasts is location and time dependent.

Many approaches have been taken to evaluate the uncertainty of a given forecast. There are two types of errors: level errors and phase errors. The use of mean or median errors in this context may be misleading as wind power forecast errors are asymmetric. This is a natural consequence of wind power being non-negative and bounded by the maximum capacity of production. This is important as the associated cost to power forecast errors is also asymmetric due to different costs for up and down power regulations, which are determined by the electricity market (\cite{tsitsiklis2015pricing}).

We propose to model wind power forecast errors using parametric stochastic differential equations (SDEs) whose solution defines a stochastic process. This resultant stochastic process describes the time evolution dynamics of wind power forecast errors while capturing properties such as a correlation structure and the inherent asymmetry. Additionally, the model we propose is agnostic of the forecasting technology and serves to complement forecasting procedures by providing a data-driven stochastic forecast. Hence, we can evaluate wind power forecasts according to their real-world performance, and we can compare different forecasting technologies. Most notably, we can simulate future wind power production given a deterministic wind power forecast. Future wind power production using Monte Carlo methods, as well as the analytic form of the proposed SDE, can be used in optimal control problems involving wind power production.

Previous attempt by (\cite{mozuma}) considered stochastic wind power forecast models based on stochastic differential equations. Here, we propose an improved model featuring time derivative tracking of the forecast, time-dependent mean reversion, modified diffusion, and non-Gaussian approximations. We apply the model to Uruguayan wind power forecasts together with historical wind power production data pertaining to the year 2019. \\

The rest of the paper is organized as follows. In Section \ref{Section_2}, we introduce the main characteristics of a real data set encompassing the normalized wind power production in Uruguay during the period April-December 2019, joint with the most accurate predictions, as highlighted in our posterior analysis, performed by one out of the three sources of forecast providers. The significant steps for constructing phenomenological models of the normalized wind power production and the forecast error based on stochastic differential equations are described in Section \ref{Section_3}. The application of the Lamperti transform with unknown parameters in Section \ref{Section_4} leads to model the forecast error through a stochastic differential equation with a unit diffusion coefficient. In Section \ref{Section_5}, we write down the expressions for the likelihood functions of the forecast error in its original space and the Lamperti space. We also derive simple approximations of the likelihood functions. Section \ref{Section_5} concludes with the description of the optimization algorithm to compute approximate maximum likelihood estimates, including the case where we expand the model comprising an initial transition from the time the forecast is performed to the time of the first forecast.

In Section \ref{Section_6}, we apply our proposed numerical estimation procedures to the Uruguay wind and forecast dataset, comparing two alternative models and assessing, for the best candidate model, the performance of three different forecast providers. Section \ref{Section_7} concludes the paper. 
The proofs of the existence, strong uniqueness, and boundedness of the SDE solutions used to model normalized wind power production and its forecast error are given in the Appendix.

@@

A view of the global discrepancy between the real production and the forecasted production, during the nine months observation period, is summarized through the forecast error histograms in Figure \ref{fig:data_curtailing}, where we also partitioned the forecast errors according to three contiguous categories of normalized generated power. Low normalized generated power corresponds to the range $[0,0.3]$, mid-power refers to the range $]0.3,0.6]$, and high-power to the range $]0.6,1]$.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_6.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_6.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_6.eps}
\caption{Wind production forecast error histograms during the period April-December 2019: low-power (upper-left plot), mid-power (upper-right plot), high-power (lower-left plot), and the global range of power (lower-right plot).}
  \label{fig:data_curtailing}
\end{figure}

We may observe that all the histograms in Figure \ref{fig:data_curtailing} exhibit skewed patterns, to a different extent, as well as extreme observations. The presence of these features can be partly explained. The data analysis highlighted that, during several 24-hour segments, the system operators decided to reduce or even cease the wind power production. Indeed, as recalled in (\cite{irena2}, p.3), ``Uruguay experiences high curtailment levels because generation exceeds demand.'' Despite the large country's interconnection capacity with Argentina and Brazil, there is no active cross-border market; the energy is traded via ad hoc short-term agreements. ``Even with interconnection capacity exceeding peak demand, the power system experiences high VRE curtailment, mostly at night when wind generation exceeds demand.''

The curtailment of the wind power production imposed by the system operators has a strong influence on the forecast error. To build a model that, driven by the available forecast, allows the inclusion of true power production with a prescribed degree of uncertainty, it is necessary to remove the data segments affected by wind curtailment.

@@

Let $(p_t)_{t \in [0,T]}$ be the available prediction function for the normalized wind power, which is an input to this approach. To start with the model specification, first, we introduce a time-dependent drift function that features the mean-reverting property as well as derivative tracking:\begin{equation}
a(X_t; p_t, \dot{p}_t, \bm{\theta}) = \dot{p}_t  - \theta_t (X_t - p_t),  \label{drift:meanrev-derivtrack}
\end{equation} 
where $ (\theta_t)_{t \in [0,T]} $ is a positive deterministic function, whose range depends on $\bm{\theta}$, as will be explained shortly.

Now, looking at the normalized wind power generation forecast process $X_t$, modeled as solution to the It\^{o} stochastic differential equation (\ref{eq:main}) with the drift specified in (\ref{drift:meanrev-derivtrack}), it is straightforward to check that $\mathbb{E} \left[X_t\right] = p_t$, given $  \mathbb{E}\left[X_0\right] = p_0$. The application of  It\^{o}'s lemma % on $g(X_t, t) = X_t e^{\int_{0}^{t} \theta_s \dif s}$, leads to 
leads to
%\begin{equation*}
%\dif \left( X_t e^{\int_{0}^{t} \theta_s \dif s} \right) = e^{\int_{0}^{t} \theta_s \dif s} (\dot{p}_t + \theta_t p_t ) \dif t  + e^{\int_{0}^{t} \theta_s \dif s}b (X_t; p_t, \dot{p}_t, \bm{\theta} ) \dif W_t,
%\end{equation*}
%whose integral form is
\begin{equation}
e^{\int_{0}^{t} \theta_s \dif s}  X_t - X_0 = \int_{0}^{t}  (\dot{p}_s + \theta_s p_s ) e^{\int_{0}^{s} \theta_u \dif u} \dif s + \int_{0}^{t}  b (X_s; p_s, \dot{p}_s, \bm{\theta} )  e^{\int_{0}^{s} \theta_u \dif u}\dif  W_s.   \label{eq:tomeanX}
\end{equation} 
Taking expectation on (\ref{eq:tomeanX}) we obtain
%\begin{align}
%\mathbb{E} \left[X_t\right] & = e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E} \left[X_0\right] +   \int_{0}^{t}  (\dot{p}_s + \theta_s p_s ) e^{\int_{0}^{s} \theta_u \dif u} \dif s   \right) \nonumber \\ 
%& =  e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E} \left[X_0\right] +   \int_{0}^{t}  \dot{p}_s e^{\int_{0}^{s} \theta_u \dif u} \dif s +  \big[ p_s \, e^{\int_{0}^{s} \theta_u \dif u} \big]^{t}_0  - \int_{0}^{t}  \dot{p}_s e^{\int_{0}^{s} \theta_u \dif u} \dif s  \right) \nonumber \\
%& =  e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E}\left[ X_0\right] + p_t \, e^{ \int_{0}^{t} \theta_s \dif s} - p_0 \right) = p_t. \label{eq:meanX}
%\end{align}
\begin{equation}
\mathbb{E} \left[X_t\right] =  e^{ - \int_{0}^{t} \theta_s \dif s} \left(  \mathbb{E}\left[ X_0\right] + p_t \, e^{ \int_{0}^{t} \theta_s \dif s} - p_0 \right) = p_t.
\label{eq:meanX}
\end{equation}

 At this stage, the process defined by (\ref{eq:main}) with drift (\ref{drift:meanrev-derivtrack}) satisfies the two following properties: 
\begin{itemize}
\item it reverts to its mean $p_t$, with a time-varying speed $ \theta_t$ that is proportional to the deviation of the process $X_t$ from its mean,
\item it tracks the time derivative $\dot{p}_t$.  
\end{itemize}

@@

\begin{figure}[H]
\centering
  \includegraphics[width=0.45\textwidth]{plots/31.eps}\includegraphics[width=0.45\textwidth]{plots/31_2.eps}
  \caption{Pointwise confidence bands fitted, for the same daily segment, through diffusion models without derivative tracking (plot on the left) and with derivative tracking (plot on the right).}
  \label{fig:derivative_tracking}
\end{figure}

The forecast and production wind power data of Uruguay are normalized with respect to the installed power capacity during the period of observation. Thus, the mean-reverting level lies in $[0,1]$, and the process $X_t$  must take values in the same interval, a requirement that is not automatically fulfilled through the derivative tracking. To impose that the state space of $X_t$ is $[0,1]$, we may choose a convenient diffusion term, and require that the time-varying parameter $ \theta_t$ satisfies an ad-hoc condition.

Let $\bm{\theta} = (\theta_0,\alpha)$, and choose a state-dependent diffusion term that avoids the process exiting from the range $[0,1]$ as follows:
  \begin{equation}
    b (X_t; \bm{\theta} )= \sqrt{2 \alpha \theta_0 X_t (1 - X_t)}
  \end{equation}
  where $\alpha >0$ is an unknown parameter that controls the path variability. This diffusion term belongs to the Pearson diffusion family and, in particular, it defines a Jacobi type diffusion. It is useful to recall that (\cite[440]{foso}) a Pearson diffusion is a stationary solution to a stochastic differential equation of the form \begin{equation}
    \dif X_t = - \theta (X_t - \mu)\dif t + \sqrt{2 \theta \left(a X_t^2 + b X_t + c\right)} \dif W_t
    \label{eq:poly_diff}
  \end{equation}
where $\theta>0$, and $a$, $b$, and $c$ are parameters such that the square root is well defined when $X_t$ is in the state space. These parameters, together with $\mu$, the mean of the invariant distribution, determine the state space of the diffusion as well as the shape of the invariant distribution.

@@

Our model (\ref{VtSDE}) for the forecast error has a diffusion term that depends on the state variable $V_t$. Under the conditions that permit to use It\^{o}'s formula on a well-chosen transformation of the process $V$, John Lamperti (\cite{lamp}) first showed that the transformed process is again a diffusion that is solution to a SDE with unit coefficient for the diffusion term. 
A vast literature refers nowadays to this result as the so-called Lamperti transform (see, for example, \cite[40--41]{iacus1}, \cite{moma}, \cite[199--203]{pani}, \cite[98--100]{saso}), which is a basic tool to obtain a SDE for the transformed process whose diffusion term does not depend anymore on the state variable. A remarkable effect of removing the state dependency from the random noise term is to increase the numerical stability of the simulated paths of the transformed process. 
For this reason, some estimation methods of the unknown parameters of non-linear SDE models incorporated the Lamperti's change of variable as part of a more complex approximation procedure (for example, in the case of one-dimensional diffusions, the local linearization method in  \cite{shoz}, or the expansion method in \cite{ait}, later extended to time-inhomogeneous SDEs in \cite{eglix}).
% later extended to multivariate diffusion in \cite{ait2}).

We consider the following Lamperti transform with unknown parameters
%{\color{red} (Remind that continuous-time models with different diffusion term are singular and the likelihood cannot be written down. Besides, it can be unstable to work with low frequency. If the frequency of the data is too low, the Lamperti transform helps to detect this problem (in principle, it is not clear how to determine the right frequency for estimation. )}
%\begin{equation}
%Z_t = h(V_t, t) = \frac{1}{\sqrt{2 \theta_0 \alpha}} \int_{\xi}^{x} \frac{1}{\sqrt{(u + p_t)(1 - u - p_t)}} \,du \Bigg{\vert}_{x = V_t} \,,
%\end{equation}
%where $\xi$ is an arbitrary point of the state space of the process $V$. The choice of $\xi = \frac{1}{2} - p_t$, where $p_t$ is a known input, 
\begin{align}
Z_t = h(V_t, t; \bm{\theta} )  & = \frac{1}{\sqrt{2 \alpha \theta_0 }} \int \frac{1}{\sqrt{(v + p_t)(1 - v - p_t)}} \dif v \Bigg{\vert}_{v = V_t}   = - \sqrt{ \frac{2}{ \alpha \theta_0 }} \arcsin (\sqrt{ 1 - V_t - p_t}) \label{eq:LampZ}
\end{align}
that, after applying It\^{o}'s formula on $h(V_t, t; \bm{\theta} )$, leads to the following SDE with state independent unit diffusion term
\begin{align}
\dif Z_t = \Bigg[  & \frac{\dot{p}_t}{ \sqrt{2 \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}}  + \frac{- \theta_t V_t}{ \sqrt{2  \alpha \theta_0 (V_t + p_t)(1 - V_t - p_t)}} - \frac{1}{4} \frac{\sqrt{2 \alpha \theta_0} \left( 1 - 2 (V_t + p_t)\right)}{\sqrt{(V_t + p_t)(1 - V_t - p_t)}}  \Bigg] \dif t + \dif W_t \,. \label{eq:stindepSDE}
\end{align}
After replacing $V_t = 1 - p_t - \sin^2 \left(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \right) $ in (\ref{eq:stindepSDE}), we obtain that the  process $Z_t$ satisfies the SDE
\begin{align}
\dif Z_t & = \left[ \frac{ \dot{p}_t  - \theta_t  \left(1 - p_t - \sin^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\sqrt{2 \alpha \theta_0} \cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)}   - \frac{1}{4}  \frac{\sqrt{2 \alpha \theta_0} \left(1 - 2 \cos^2 \Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big) \right) }{\cos\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)  \sin\Big(- \sqrt{ \frac{ \alpha \theta_0}{2} } Z_t \Big)} \right]\dif t + \dif W_t \nonumber \\
&  = \left[  \frac{  2  \dot{p}_t - \theta_t (1 - 2 p_t)  + (\alpha \theta_0 - \theta_t) \cos(- \sqrt{2 \alpha \theta_0 } Z_t) }{\sqrt{2 \alpha \theta_0} \sin{(- \sqrt{2 \alpha \theta_0} Z_t)}}  \right] \dif t + \dif W_t.  \label{eq:stindepSDE2}
\end{align}

To summarize the effect of the Lamperti transform visually, we can see in the next Figure (\ref{fig:LP_transitions}) how the forecast error transition histograms (without curtailment) modify in comparison with Figure (\ref{fig:error_transitions}).

In this case, the Lamperti transform has been applied using the optimal estimates of the parameters in the SDE model (\ref{VtSDE}), obtained applying our numerical procedure detailed later in Subsection \ref{opt_sec}.

\begin{figure}[H]
\centering
\includegraphics[width=0.45\textwidth]{plots/LP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/MP_t_LP.eps}\\
\quad\\
\includegraphics[width=0.45\textwidth]{plots/HP_t_LP.eps}
\includegraphics[width=0.45\textwidth]{plots/AP_t_LP.eps}
\caption{Lamperti transformed forecast error transition histograms during the period April-December 2019 without wind power production curtailment: low-power (upper-left plot), mid-power (upper-right plot), high-power (lower-left plot), and the global range of power (lower-right plot).}
  \label{fig:LP_transitions}
\end{figure}

The shape of the forecast error transition histograms after applying the Lamperti transform has much resemblance with a Gaussian distribution, motivating toward the use of Gaussian-like approximations of the unknown density transition functions of the process $Z_t$.

{\color{red}Remark: Moreover, this obtained Gaussian distribution supports the validity of the choice of our model diffusion coefficient given by (\ref{eq:poly_diff}).}

@@

\item Use the training dataset to find the SDE's first and second moments as explained in Subsection \ref{moments_ODEs}.
\item Match the proxy distribution moments with the SDE's moments.
\item Evaluate the negative log-likelihood using the training dataset.
